NMRlipids Databank: Overlay Databank of Lipid Membrane Simulations Arising from Open Collaboration

### 2022-09-07 SI.tex
Update manuscript

@@

>>>>

\makeatletter
\newcommand*{\centerfloat}{%
  \parindent \z@
  \leftskip \z@ \@plus 1fil \@minus \textwidth
  \rightskip\leftskip
  \parfillskip \z@skip}
\makeatother

@@

\pagebreak
\section{Correlations between area per lipid and thickness with order parameters and form factors}

\begin{figure}[!h]
    \centerfloat
    \includegraphics[width = 180mm]{Figures/QualityCorrelationsSI.pdf}
    \caption{Dependence of form factor, electron density profiles along membrane normal, and C-H bond orderparameters (from top to bottom) on simulation box size with different cholesterol concentration. }
    \label{fig:sizedependence}
\end{figure}

\pagebreak
\section{Dependence of form factor and order parameters in the size of simulation}

\begin{figure}[!h]
    \centerfloat
    \includegraphics[width = 240mm]{Figures/SizeDependence.pdf}
    \caption{Dependence of form factor, electron density profiles along membrane normal, and C-H bond orderparameters (from top to bottom) on simulation box size with different cholesterol concentration. }
    \label{fig:sizedependence}
\end{figure}

### 2022-09-07 main.tex
Update manuscript

\begin{figure}[tbp]
    \centering
    \includegraphics[width=180mm]{Figures/quality.pdf}
    \caption{a) Correlations between membrane area per lipid, thickness, second minima of form factor and average order parameter of the {\it sn}-1 acyl chain extracted from the NMRlipids databank. All Pearson correlation coefficients shown in bottom left corner have p-value below 0.001.
    b) Evaluation against experimental data exemplified for the simulation giving the best qualities in overall, for headgroup and x-ray scattering form factor.
    c) Change in area per lipid upon addition of POPE to POPC in different simulation extracted from the NMRlipids databank and quality evaluation table of POPE simulations.
    d) Quality evaluation table showing the best 13 simulations according to the overall order parameter quality.}
    \label{fig:quality}
\end{figure}

@@

The quality of conformational ensembles of individual lipid molecules in simulations are evaluated using the C-H bond order parameters which can be directly compared with robust experimental data~\cite{ollila16}.
%The C--H bond order parameters were calculated directly
%from the carbon and hydrogen positions using the 
%definition
The order parameters are defined as
\begin{equation}\label{OP}
S_{\rm CH}=\frac{1}{2}\left\langle 3\cos^2\theta -1 \right\rangle,
\end{equation}
where $\theta$ is the angle between the C--H bond and the membrane normal
%(taken to align with $z$, with bilayer periodicity in the $xy$-plane).
and angular brackets denote the ensemble average, i.e. average over all sampled configurations of all lipids in a simulation.
%by estimating
%the average probability of C-H bond order parameters to agree with the values from NMR experiments. 
%As a first step, the quality of each order parameter is evaluated using the negative logarithm of probability to find the simulated order parameter within the experimental error bars
%\begin{equation}
%    S_q = -\log_{10}(P),    
%\end{equation}
Because conformational ensembles of individual lipids are independent in a simulation of a fluid lipid bilayer, the number of independent sample points for each C-H bond, $n$, equals the number of lipids in a simulation. Denoting the $S_{\rm CH}$ from Eq.~\ref{OP} as the sample mean and $s$ its variance calculated over individual lipids, 
%the probability the true mean of the order parameter is given by the %Student's t-distribution 
%Because each C-H bond order parameter is calculated as an average over individual lipids in a simulation
%\begin{equation}\label{studentT}
$\frac{S_{\rm CH}-\mu}{s/\sqrt{n}}$
%\end{equation}
%where $\bar{X}$ is the sample mean (order parameter calculated from simulations),
%$s$ is the sample variance (variance of order parameters in different lipids)
%and 
%has a 
has a Student's t-distribution with $n-1$ degrees of freedom, where $\mu$ is the real mean of the order parameter. 
%shifted by $S_{\rm CH}$ and scaled by $s/\sqrt{n}$.
Because lipid bilayer simulations contain at least dozens of lipids, the Student's t-distribution could be safely approximated with a normal distribution. However, when calculating the probability for a simulation to locate within experimental order parameters, the normal distribution gives values below the numerical accuracy of computers for simulations values far from experiments. To avoid such numerical instability, we use the first order Student's t-distribution.% for the equation \ref{studentT}.
Therefore, we estimate the probability for the simulation order parameter to locate within experimental error bars from equation
\begin{equation}\label{probability}
  P = f \left( \frac{S_{\rm CH} - (S_{\rm exp}+\Delta S_{\rm exp})}{s/\sqrt{n}} \right) - f \left( \frac{S_{\rm CH} - (S_{\rm exp}-\Delta S_{\rm exp})}{s/\sqrt{n}} \right),
\end{equation}
where f(t) is the first order Student's t-distribution.
% shifted to $\bar{X}$ and scaled with $s$.

@@

Qualities of form factors were evaluated using the same approach as in SIMtoEXP program~\cite{kucerka10}. Because experiments give form factors only in relative intensity scale, they were scaled to the simulation data using equation
\begin{equation}
    k_e = \frac{\sum_{i=1}^{N_q} \frac{|F_s(q_i)||F_e(q_i)|}{(\Delta F_e(q_i))^2}}{\sum_{i=1}^{N_q} \frac{|F_e(q_i)|^2}{(\Delta F_e(q_i))^2}}.
\end{equation}
The quality of each form factor was then calculated from the equation
\begin{equation}
    \chi^2 = \frac{\sqrt{\sum_{i=1}^{N_q}(|F_s(q_i)|-k_e|F_e(q_i)|)^2/(\Delta F_e(q_i))^2}}{\sqrt{N_q-1}},
\end{equation}
where $F_s$ and $F_e$ are form factors from a simulation and experiment, respectively, and summation goes over the experimentally available $N_q$ points.

>>>>

\begin{figure}[tbp]
    \centering
    \includegraphics[width=180mm]{Figures/quality3.pdf}
    \caption{a) Correlations between membrane area per lipid, thickness, second minima of form factor and average order parameter of the {\it sn}-1 acyl chain extracted from the NMRlipids databank. All Pearson correlation coefficients shown in bottom left corner have p-value below 0.001.
    b) Evaluation against experimental data exemplified for the simulation giving the best qualities in overall, for headgroup and x-ray scattering form factor.
    c) Change in area per lipid upon addition of POPE to POPC in different simulation extracted from the NMRlipids databank and quality evaluation table of POPE simulations.
    d) Quality evaluation table showing the best 13 simulations according to the overall order parameter quality.}
    \label{fig:quality}
\end{figure}

@@

\subsubsection{Conformational ensembles using C-H bond order parameters}
The quality of conformational ensembles of individual lipid molecules in simulations are evaluated using the C-H bond order parameters which can be directly compared with robust experimental data~\cite{ollila16}.
%The C--H bond order parameters were calculated directly
%from the carbon and hydrogen positions using the 
%definition
The order parameters are defined as
\begin{equation}\label{OP}
S_{\rm CH}=\frac{1}{2}\left\langle 3\cos^2\theta -1 \right\rangle,
\end{equation}
where $\theta$ is the angle between the C--H bond and the membrane normal
%(taken to align with $z$, with bilayer periodicity in the $xy$-plane).
and angular brackets denote the ensemble average, i.e. average over all sampled configurations of all lipids in a simulation.
%by estimating
%the average probability of C-H bond order parameters to agree with the values from NMR experiments. 
%As a first step, the quality of each order parameter is evaluated using the negative logarithm of probability to find the simulated order parameter within the experimental error bars
%\begin{equation}
%    S_q = -\log_{10}(P),    
%\end{equation}
Because conformational ensembles of individual lipids are independent in a simulation of a fluid lipid bilayer, the number of independent sample points for each C-H bond, $n$, equals the number of lipids in a simulation. Denoting the $S_{\rm CH}$ from Eq.~\ref{OP} as the sample mean and $s$ its variance calculated over individual lipids, 
%the probability the true mean of the order parameter is given by the %Student's t-distribution 
%Because each C-H bond order parameter is calculated as an average over individual lipids in a simulation
%\begin{equation}\label{studentT}
$\frac{S_{\rm CH}-\mu}{s/\sqrt{n}}$
%\end{equation}
%where $\bar{X}$ is the sample mean (order parameter calculated from simulations),
%$s$ is the sample variance (variance of order parameters in different lipids)
%and 
%has a 
has a Student's t-distribution with $n-1$ degrees of freedom, where $\mu$ is the real mean of the order parameter. 
%shifted by $S_{\rm CH}$ and scaled by $s/\sqrt{n}$.
Because lipid bilayer simulations contain at least dozens of lipids, the Student's t-distribution could be safely approximated with a normal distribution. However, when calculating the probability for a simulation to locate within experimental order parameters, the normal distribution gives values below the numerical accuracy of computers for simulations values far from experiments. To avoid such numerical instability, we use the first order Student's t-distribution.% for the equation \ref{studentT}.
Therefore, we estimate the probability for the simulation order parameter to locate within experimental error bars from equation
\begin{equation}\label{probability}
  P = f \left( \frac{S_{\rm CH} - (S_{\rm exp}+\Delta S_{\rm exp})}{s/\sqrt{n}} \right) - f \left( \frac{S_{\rm CH} - (S_{\rm exp}-\Delta S_{\rm exp})}{s/\sqrt{n}} \right),
\end{equation}
where f(t) is the first order Student's t-distribution.
% shifted to $\bar{X}$ and scaled with $s$.

@@

\subsubsection{Membrane dimensions with x-ray scattering form factors}
While C-H bond order parameters relate to the conformational ensembles of individual lipids,
the x-ray scattering form factors depend on membrane dimensions and density distribution~\cite{ollila16}.
Because experiments give form factors only in relative intensity scale, they are typically scaled to the simulation data.
Here we use the same scaling method as in SIMtoEXP program~\cite{kucerka10} where experimental form factor intensities
are scaled by a factor defined as
\begin{equation}
    k_e = \frac{\sum_{i=1}^{N_q} \frac{|F_s(q_i)||F_e(q_i)|}{(\Delta F_e(q_i))^2}}{\sum_{i=1}^{N_q} \frac{|F_e(q_i)|^2}{(\Delta F_e(q_i))^2}},
\end{equation}
%The quality of each form factor was then calculated from the equation
%\begin{equation}
%    \chi^2 = \frac{\sqrt{\sum_{i=1}^{N_q}(|F_s(q_i)|-k_e|F_e(q_i)|)^2/(\Delta F_e(q_i))^2}}{\sqrt{N_q-1}},
%\end{equation}
where $F_s$ and $F_e$ are form factors from a simulation and experiment, respectively, and summation goes over the experimentally available $N_q$ points. 
Quality measure based on differences in simulated and experimental form factors accross available q-range is also defined in the SIMtoEXP program.
However, as shown in Fig.~\ref{??}, the lobe heigths in the simulated form factors depend on the simulation box size,
but locations of minima does not. Therefore, quality measure depending on lobe heights is also dependent on simulation box size.
In order to define a quality measure which is independent on simulation box size, we measure the form factor quality based on the
location of the first minima observed in experiments. As shown in Fig.~\ref{??}, this minima correlates well with the thickness of a membrane.
Although the correlation in second minima would be even stronger, we have chosen to use the first minima because automatic definition
of second minima is inaccurate in some experimental data. In practise, we first filter the fluctuations from the form factor data using
Savitzky-Golay filter (window lenght 30 and polynomial order 1) and locate the first minima above 0.1\,$AA^2$ from both simulation and experimental data.
The quality of a form factor is then defined as Eucledian distance between the minima in simulated and experimental form factors, $FF_q = |FF_{\rm{min}}^{\rm{sim}}-FF_{\rm{min}}^{\rm{exp}}|$.
